Biogeography and Environmental Conditions Shape Phage and Bacteria Interaction Networks Across the Healthy Human Microbiome

### 2017-03-26 doc/manuscript.Rmd
Some minor updates.

---
title: Biogeography and Environmental Conditions Shape Phage and Bacteria Interaction Networks Across the Healthy Human Microbiome
author: 
- name: Geoffrey D Hannigan
  affiliation: 1
- name: Melissa B Duhaime
  affiliation: 2
- name: Danai Koutra
  affiliation: 3
- name: Patrick D Schloss
  affiliation: 1,\*
address:
  - code: 1
    address: Department of Microbiology & Immunology, University of Michigan, Ann Arbor, Michigan, 48108
  - code: 2
    address: Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, Michigan, 48108
  - code: 3
    address: Department of Computer Science, University of Michigan, Ann Arbor, Michigan, 48108
  - code: \*
    address: To whom correspondence may be addressed.
output:
  md_document:
    variant: markdown
fontsize: 11pt
mainfont: 'Helvetica'
---

@@

# Abstract
Viruses and bacteria are critical components to the human microbiome and play important roles in health and disease. Previous work has relied on studying bacteria and phages in isolation (e.g. taxonomic classification, alpha diversity, and beta diversity), reducing them to two separate communities. This approach cannot capture how these communities interact to share information (e.g. horizontal gene transfer) and maintain stability. We developed and implemented a network-based analytical approach to provide an initial understanding of phage-bacteria network diversity across the human body and over time. We built a machine learning algorithm to predict which phages will infect which bacteria in each microbiome. This model was applied to paired viral and bacterial metagenomic sequence sets from three previously published human cohorts. We organized the predicted interactions into networks that allowed us to evaluate the diversity of phage-bacteria connectedness and network structure across the human body. We found that gut and skin network structure was person-specific, and was only weakly conserved among cohabitating family members. High fat diets and obesity were associated with less connected networks. There were significant differences in network structure between skin sites, with those more exposed to the external environment being less connected and more prone to instability. This study characterizes the diversity of microbiome networks across the human body, and provides a baseline for future studies to investigate the role of ecological networks in disease states.

**Word Count**: 231 / 250

@@

Viruses and bacteria are critical components to the human microbiome and play important roles in health and disease. Bacterial communities have been associated with diseases including a wide range of skin conditions [@Hannigan:2013im], acute and chronic wound healing conditions [@Hannigan:2014be; @Loesche:2016ev], and gastrointestinal diseases including inflammatory bowel disease [@He:2016ch; @Norman:2015kb], *Clostridium difficile* infections [@Seekatz:2016fz], and colorectal cancer [@Zackular:2014fba; @Baxter:2014hb]. Altered viromes (virus communities consisting primarily of bacteriophages in humans) have also been associated with various diseases and environmental perturbations including inflammatory bowel disease [@Norman:2015kb; @Manrique:2016dx], periodontal disease [@Ly:2014ew], and others [@Monaco:2016ita; @Hannigan:2015fz; @Minot:2011ez; @SantiagoRodriguez:2015gd; @Abeles:2015dy; @Abeles:2014kj]. Individual phages within these communities are capable of lysing their bacterial hosts and modulating their functionality through horizontal gene transfer (i.e. transduction) and altering bacterial host gene expression. At the community level, the human virome has begun to be implicated in promoting antibiotic resistance throughout human-associated microbial communities [@Modi:2013fia; @Hannigan:2015fz]. These virus community shifts are not reflections of the bacterial communities, but rather act in concert with bacteria as one single overall community [@Norman:2015kb; @Haerter:2014ii].

Previous work has relied on studying bacteria and phages in isolation (e.g. taxonomic classification, alpha diversity, and beta diversity), reducing them to two separate communities. In reality, bacteria and phage communities are dynamic and complex, acting in concert to share genetic information (e.g. horizontal gene transfer) and to maintain stable ecosystems, and removal of members can disrupt or even collapse those ecosystems [@Haerter:2014ii; @Harcombe:2005fd; @Middelboe:2001fl; @Poisot:2011jc; @Thompson:2012ki; @Moebus:1981kp; @Flores:2013hc; @Poisot:2012fh; @Poisot071993; @Flores:2011bh; @Jover:2015ev]. Relationship-based network approaches allow us to capture this information, including community resilience to instability and the potential for information flow. The conventional approach to understanding the human microbiome is unable to capture how these communities interact and influence each other. We bridge this knowledge gap by characterizing the human bacterial and phage communities by their *relationships*, leveraging machine learning and graph theory techniques. We focus on characterizing the network diversity of the healthy human microbiome (including the skin and gut), so as to provide a foundation on which we and others can build further studies into the network dynamics associated with disease states.

To begin characterizing the networks of bacteria and phages within the human microbiome, we leveraged three published microbiome datasets with paired virus and bacterial metagenomic sequence sets [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei]. These datasets included gut and skin samples, allowing us to gain broader insights into human microbiome network diversity across different body sites. Our approach built off of previous large-scale phage-bacteria microbiome network analyses by inferring interactions using metagenomic datasets, instead of using culture-based techniques [@Moebus:1981kp; @Flores:2013hc]. Our metagenomic interaction inference model is powered beyond previous models by its inclusion of protein interaction data, inclusion of negative interactions as well as positive, and the use of a more sophisticated machine learning algorithm instead of the linear models used by some groups [@Edwards:2015iz].

Using this approach, we were able to go beyond contemporary isolated methods to uncover a basic understanding of the community network dynamics associated with healthy human phage and bacteria. By building and utilizing a microbiome network, we showed that different people, body sites, and anatomical locations not only support distinct microbiome membership and diversity [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei; @Grice:2009eea; @Findley:2013jf; @Costello:2009im, @Consortium:2012iz], but also support communities with distinct communication structures and propensities toward community instability. By understanding the healthy state of network structures across the human body, we empower future studies to begin investigating how these community structures change in disease or otherwise altered states.

@@

## Establishing a Model of Phage-Bacteria Infectious Networks
We studied the differences in microbiome phage-bacteria interaction networks across healthy human bodies by leveraging previously published sequence sets containing purified virome samples paired with bacterial metagenomes from whole metagenomic shotgun sequences. Our study contained three datasets, including a study of the impact of diet on the healthy human gut virome [@Minot:2011ez], the impact of anatomical location on the healthy human skin virome [@Hannigan:2015fz], and the viromes of monozygotic twins and their mothers [@Reyes:2010cwa; @Turnbaugh:2009ei]. The viromes associated with these datasets were subjected to virus-like particle (VLP) purification to eliminate other organism DNA including bacteria, fungi, and humans, and allowed us to assess the actively replicating virome.

Bacterial and viral sequences were quality filtered and assembled into contigs. The published datasets we used were conducted over the span of five years using different methods and technologies, and therefore yielded different sequence abundances **(Supplemental Figure \ref{SequenceStats} A-B)**. Because contig assembly most commonly returns genome fragments, we clustered related bacteria and phage contigs by k-mer frequency and co-abundance using the CONCOCT algorithm. These clusters represent operationally defined units of related bacteria and phage genomes that we defined as operational genomic units (OGUs). These OGUs are conceptually similar to the operational taxonomic unit (OTU) and operational protein family (OPF) definitions used for grouping highly similar 16S rRNA gene and open reading frame sequences, respectively [@Schloss:2008ht]. The resulting contigs and OGUs demonstrated high sequence coverage and length **(Supplemental Figure \ref{ContigStats} - \ref{ClusterStats})**.

We predicted which phage OGUs infected which bacteria using a random forest model trained on experimentally validated infectious relationships from six previous publications [@Jensen:1998vh;@Malki:2015tm;@Schwarzer:2012ez;@Kim:2012dh;@Matsuzaki:1992gw;@Edwards:2015iz]. This training set contained diverse bacteria and phages, with both broad and specific infectious ranges **(Supplemental Figure \ref{ValidationOverview} A - B)**. Phages with linear and circular genomes, as well as ssDNA and dsDNA genomes, were included in the analysis. Because this was a DNA sequencing study, RNA phages were considered in the analysis **(Supplemental Figure \ref{ValidationOverview} C-D)**. This training set included both positive relationships (a phage infects a bacterium) and negative relationships (a phage does not infect a bacterium). This built on previous work, which focused only on positive relationships, by allowing us to validate the false positive *and* false negative rates associated with our candidate models.

Four phage and bacterial genomic markers were used to predict infectious relationships between bacteria and phages: 1) genome nucleotide similarities, 2) gene amino acid sequence similarities, 3) CRISPR targeting of phages by bacterial CRISPR spacer sequences, and 4) similarity of protein families known to be associated with known protein-protein interactions [@Orchard:2014hq]. The resulting random forest model exhibited high performance with an AUC of `r signif(genmodelper$ROC, digits = sigfig)`, a sensitivity of `r signif(genmodelper$Sens, digits = sigfig)`, and a specificity of `r signif(genmodelper$Spec, digits = sigfig)` **(Figure \ref{RocCurve} A)**. The most important predictor in the model was nucleotide similarity between genes, followed by nucleotide similarity of whole genomes **(Figure \ref{RocCurve} B)**. Protein family interactions were moderately important to the model, while CRISPRs were minimally important, likely because they were redundant with the blast information. Approximately one third of the relationships yielded no score and were automatically classified as being non-infectious **(Figure \ref{RocCurve} C)**.

We used our random forest model to classify the relationships between bacteria and phages in the experimental datasets. The relationships within the three studies were used to construct one master network, containing the three study sub-networks, which themselves each contain sub-networks for each sample **(Figure \ref{RocCurve} D)**. Metadata including study, sample ID, disease, and abundance within the community (based on sequence count) were also stored in the multi-study master network to allow for effective parsing and downstream analysis **(Supplemental Figure \ref{NetworkDiagram})**. The resulting master network was highly connected and contained `r prettyNum(genfigurestats[c(genfigurestats$class %in% "Total"),"Edges"], big.mark = ",")` infectious relationships among `r prettyNum(genfigurestats[c(genfigurestats$class %in% "Total"),"Vertices"], big.mark = ",")` nodes, `r prettyNum(nestats[c(nestats$cats %in% "PhageNodes"),"values"], big.mark = ",")` of which represented phages and `r prettyNum(nestats[c(nestats$cats %in% "BacteriaNodes"),"values"], big.mark = ",")` that represented bacteria. Although the network was highly connected, not all relationships were present in all samples. Furthermore, relationships were weighted by relative abundance of their associated bacteria and phage, meaning that lowly abundant relationships could be present but insignificant compared to those that were more highly abundant. Like the master network, the skin network exhibited a diameter of `r prettyNum(genfigurestats[c(genfigurestats$class %in% "SkinStudy"),"Diameter"], big.mark = ",")` (measure of graph size; the greatest number of traversed vertices required between two vertices) and included almost all of the master network nodes and edges **(Figure \ref{RocCurve} E - F)**. The gut diet and twin sample sets each contained less than 150 vertices, less than 20,000 relationships, and diameters of `r prettyNum(genfigurestats[c(genfigurestats$class %in% "DietStudy"),"Diameter"], big.mark = ",")`, suggesting more sparsely related phages and bacteria **(Figure \ref{RocCurve} E - F)**.

## Individuality of Microbial Networks
Previous work has show community membership and diversity are highly personal, with signatures remaining more similar to themselves over time compared to other people [@Grice:2009ee; @Hannigan:2015fz; @Minot:2013ih]. Using our microbiome network model, we determined whether this personal conservation extended to network structure. We calculated the degree of dissimilarity between each subject's network, based on phage and bacteria abundance, as well as centrality. Centrality was evaluated by first calculating the weighted eigenvector centrality of all bacteria and phages within each sample graph. Conceptually, this metric defines central phages as those that are highly abundant and infect many bacteria which themselves are abundant and infected by many other phages. Bacterial centrality was defined in the same way. We then calculated the similarity of community networks using the overall weighted eigenvector centrality of all nodes between all samples. Biologically, samples with similar network structures are interpreted as having similar capacities to influence other microbes within the community, transmit genetic material (e.g. horizontal gene transfer), and maintain stability.

We found that gut microbiome network structure was highly individual specific, with networks clustering significantly by person (anosim p-value = `r signif(interstats[c(interstats$site %in% "DietAnosim"), "prob"], digits = sigfig)`, anosim statistic R = `r signif(interstats[c(interstats$site %in% "DietAnosim"), "abserr"], digits = sigfig)`, **Figure \ref{intradiv} A)**. Network dissimilarity within people over the 8-10 day sampling period was significantly less than the average dissimilarity between that person and others **(Figure \ref{intradiv} B)**. We evaluated the significance of an individual's network being more similar to itself than others by fitting a probability distribution to the reductions in intrapersonal dissimilarity compared to interpersonal dissimilarity (Lines between points in Figure \ref{intradiv} B - D). By taking the integral of the probability distribution from negative infinity to zero, we were able to calculate the probability that an individual's network structure was more similar to itself than others, in a given population. Thus a probability of 1 would mean that all members in a given population will be expected to be more similar to themselves compared to others, and 0.5 means they are no more likely to be more similar to themselves than dissimilar. This is a Bayesian approach that tells us how prevalent we expect this trend to be. The probability of intrapersonal diversity of an individual being less than interpersonal diversity in our dataset was `r signif(interstats[c(interstats$site %in% "Diet"), "prob"], digits = sigfig)` (absolute error = `r signif(interstats[c(interstats$site %in% "Diet"), "abserr"], digits = sigfig)`, **Figure \ref{intradiv} B)**.

Skin microbiome network structure was also conserved within individuals, although not as strongly as the gut. The probability of a given skin site being more similar to itself after a month compared to a different person at the same time was `r signif(interstats[c(interstats$site %in% "Skin"), "prob"], digits = sigfig)` (absolute error = `r signif(interstats[c(interstats$site %in% "Skin"), "abserr"], digits = sigfig)`, **Figure \ref{intradiv} C)**. This distribution was similar when separated by anatomical sites, suggesting this was an accurate representation of overall skin network dissimilarity **(Supplemental Figure \ref{allskin})**.

The conservation of gut network structures was only weakly associated with families. The gut network structures were more similar between twins and their mothers (intrafamily) compared to other families of twins and mothers (inter-family) **(Figure \ref{intradiv} D)**. The probability of a family member being more similar to another family member compared to non-family members was `r signif(interstats[c(interstats$site %in% "Twins"), "prob"], digits = sigfig)` (absolute error = `r signif(interstats[c(interstats$site %in% "Twins"), "abserr"], digits = sigfig)`).

## Role of Diet & Obesity in Gut Microbiome Conectivity
Diet is a major environmental factor that influences resource availability and gut microbiome composition and diversity, including bacteria and phages [@Minot:2011ez; @Turnbaugh:2009hf; @David:2014cl]. Previous work in isolated culture-based systems has suggested that changes in nutrient availability are associated with altered phage - bacteria network structures, although this has yet to be applied to humans [@Poisot:2011jc]. Understanding the impact diet might have on gut microbiome networks is a valuable contribution to the baseline understanding we aim to achieve here. We therefore hypothesized that network structure would be altered by changes in diet. We also described a potential association between gut network structure and obesity, a disease linked to diet and potentially the microbiome [@Sze:2016iw].

We evaluated the differences in gut network structure by quantifying how central each sample's network was on average across all microbes. We accomplished this by utilizing two common centrality metrics: degree centrality and closeness centrality. **Degree centrality**, the simplest centrality metric, was defined as the number of connections each phage made to bacteria, or each bacterium made to phages. Because this metric alone offers only minimal insight, we supplemented it with measurements of closeness centrality. **Closeness centrality** is a metric of how close each phage or bacterium is to all of the other phages and bacteria in the network. A higher closeness centrality suggests that genetic information or the effects of altered abundance would be more impactful to all other microbes in the system. Networks with higher closeness centrality also indicates an overall greater degree of connections, which indicates a greater resilience to instability. Because these values are assigned to each phage and bacterium within each network, we calculated the average connectedness and corrected for the maximum potential degree of connectedness to obtain a single value for the connectedness of each graph.

We found that gut microbiome network structures associated with high fat diets were less connected than those of low fat diets **(Figure \ref{dietnetworks} A-B)**. Tests for statistical differences were not performed, so as to prevent misleading interpretations from the small sample size. High fat diets exhibited less degree centrality **(Figure \ref{dietnetworks} A)**, meaning bacteria were overall targeted by less phages and phage tropism was more specific. High fat diets also exhibited decreased closeness centrality **(Figure \ref{dietnetworks} B)**, meaning the microbes were more distant from other microbes in the community, making genetic information transfer and the impact of altered abundance less capable of impacting other bacteria and phages within the network.

In addition to diet, there was also an association between obesity and network structure **(Figure \ref{dietnetworks} C-D)**. The obesity-associated network demonstrated a higher degree centrality **(Figure \ref{dietnetworks} C)** but a less closeness centrality, compared to the healthy controls **(Figure \ref{dietnetworks} D)**. This meant that the obesity network was overall less connected, having microbes further from all other microbes within the community.

## Variation of Network Structure Across the Human Skin Landscape
Extensive work has illustrated differences in healthy human skin microbiome between anatomical sites, including bacteria, viruses, and fungi [@Grice:2009ee; @Findley:2013jf; @Hannigan:2015fz]. These communities vary by degree of skin moisture, oil, and environmental exposure. We hypothesized that like microbial composition and diversity, microbial network structure differs between anatomical sites. We addressed this hypothesis by evaluating the changes in network structure between anatomical sites within our skin dataset.

We quantified the average centrality of each sample using the weighted eigenvector centrality metric, which yields a larger value for phages and bacteria that are more abundant and have more relationships to other bacteria and phages that themselves have more relationships. We found that intermittently moist skin sites (dynamic sites that fluctuate between being moist and dry) were significantly less connected than the more stable moist and sebaceous environments (p-value < 0.001, **Figure \ref{skinnetwork} A)**. We also found that skin sites that were protected from the environment (occluded) were much more highly connected than those that were constantly exposed to the environment or only intermittently occluded (p-value < 0.001, **Figure \ref{skinnetwork} B)**.

@@

# Discussion
We developed and implemented a network-based analytical approach that allowed us to begin evaluating the basic properties of the human microbiome through bacteria and phage relationships, instead of membership or diversity of separate communities. The goal of this study was to provide an initial understanding of how phage-bacteria networks differ throughout the human body, so as to provide a baseline for future studies of how microbiome networks differ in disease states. This goal is similar to the initial studies of bacteria and viral diversity across the human body, as well as other environments [@Grice:2009eea; @Findley:2013jf; @Hannigan:2015fz; @Costello:2009im, @Consortium:2012iz; @Schloss:2005hz; @Minot:2011ez]. Our approach was advantageous over previous microbiome network analyses because it did not rely on inappropriate linear correlations of abundance, it was trained on both positive and negative relationships, and it involved utilization of network theory concepts.

>>>>

---
title: Biogeography and Environmental Conditions Shape Phage and Bacteria Interaction Networks Across the Healthy Human Microbiome
author: 
- name: Geoffrey D Hannigan
  affiliation: 1
- name: Melissa B Duhaime
  affiliation: 2
- name: Danai Koutra
  affiliation: 3
- name: Patrick D Schloss
  affiliation: 1,\*
address:
  - code: 1
    address: Department of Microbiology & Immunology, University of Michigan, Ann Arbor, Michigan, 48109
  - code: 2
    address: Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, Michigan, 48109
  - code: 3
    address: Department of Computer Science, University of Michigan, Ann Arbor, Michigan, 48109
  - code: \*
    address: To whom correspondence may be addressed.
output:
  md_document:
    variant: markdown
fontsize: 11pt
mainfont: 'Helvetica'
---

@@

# Abstract
Viruses and bacteria are critical components of the human microbiome and play important roles in health and disease. Previous work has relied on studying bacteria and phages in isolation, reducing them to two separate communities. These approaches have not captured how these communities interact to share information (e.g. horizontal gene transfer) and maintain stability. We developed and implemented a network-based analytical approach to provide an initial understanding of phage-bacteria network diversity throughout the human body. We accomplished this by building a machine learning algorithm to predict which phages infected which bacteria in a given microbiome. This algorithm was applied to paired viral and bacterial metagenomic sequence sets from three previously published human cohorts. We organized the predicted interactions into networks that allowed us to evaluate the diversity of phage-bacteria connectedness across the human body. We focused on establishing a foundational understanding of microbiome network structures across the human body. We found that gut and skin network structure was person-specific, and was not conserved among cohabitating family members. High fat diets and obesity were associated with less connected networks. There were significant differences in network structure between skin sites, with those exposed to the external environment being less connected and more prone to instability. Altogether this study characterizes the diversity of microbiome networks across the human body, and provides a baseline for future studies to investigate the role of ecological networks in disease states.

**Word Count**: 236 / 250

@@

Viruses and bacteria are critical components to the human microbiome and play important roles in health and disease. Bacterial communities have been associated with diseases including a wide range of skin conditions [@Hannigan:2013im], acute and chronic wound healing conditions [@Hannigan:2014be; @Loesche:2016ev], and gastrointestinal diseases including inflammatory bowel disease [@He:2016ch; @Norman:2015kb], *Clostridium difficile* infections [@Seekatz:2016fz], and colorectal cancer [@Zackular:2014fba; @Baxter:2014hb]. Altered viromes (virus communities consisting primarily of bacteriophages in humans) have also been associated with various diseases and environmental perturbations including inflammatory bowel disease [@Norman:2015kb; @Manrique:2016dx], periodontal disease [@Ly:2014ew], spread of antibiotic resistance [@Modi:2013fia; @Hannigan:2015fz], and others [@Monaco:2016ita; @Hannigan:2015fz; @Minot:2011ez; @SantiagoRodriguez:2015gd; @Abeles:2015dy; @Abeles:2014kj]. These shifts in virus community composition are not reflections of the bacterial communities, but rather act in concert with bacteria as one single overall community [@Norman:2015kb; @Haerter:2014ii]. Individual phages within these communities are capable of lysing their bacterial hosts and modulating their functionality through horizontal gene transfer (i.e. transduction) and altering bacterial host gene expression.

Previous work has relied on studying bacterial and phage communities in isolation, reducing them to two separate communities. In reality, bacteria and phage communities are dynamic and complex, acting in concert to share genetic information and to maintain stable ecosystems, and removal of members can disrupt or even collapse those ecosystems [@Haerter:2014ii; @Harcombe:2005fd; @Middelboe:2001fl; @Poisot:2011jc; @Thompson:2012ki; @Moebus:1981kp; @Flores:2013hc; @Poisot:2012fh; @Poisot071993; @Flores:2011bh; @Jover:2015ev]. The conventional approach to understanding the human microbiome is unable to capture how these communities interact and influence each other, but relationship-based network approaches allow us to capture this information. We aimed to bridge this knowledge gap by characterizing human bacterial and phage communities by their *relationships*, leveraging machine learning and graph theory techniques. We focused on characterizing the network diversity of the healthy human microbiome (including the skin and gut), so as to provide a foundation on which we and others can build further studies into the network dynamics associated with disease states.

We began characterizing human microbiome phage and bacteria networks by leveraged three published microbiome datasets with paired virus and bacterial metagenomic sequence sets [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei]. These datasets included gut and skin samples, allowing us to gain broader insights into human microbiome network diversity across different body sites. Our approach built off of previous large-scale phage-bacteria microbiome network analyses by inferring interactions using metagenomic datasets, instead of using culture-based techniques [@Moebus:1981kp; @Flores:2013hc]. Our metagenomic interaction inference model was powered beyond previous models by its inclusion of protein interaction data, inclusion of negative interactions as well as positive, and the use of a more sophisticated machine learning algorithm instead of linear models [@Edwards:2015iz].

Using this approach, we were able to go beyond contemporary isolated methods to provide a basic understanding of the community network dynamics associated with healthy human phages and bacteria. By building and utilizing a microbiome network, we found that different people, body sites, and anatomical locations not only support distinct microbiome membership and diversity [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei; @Grice:2009eea; @Findley:2013jf; @Costello:2009im, @Consortium:2012iz], but also support communities with distinct communication structures and propensities toward community instability. By understanding the healthy state of network structures across the human body, we will empower future studies to begin investigating how these community structures change in disease or otherwise altered states.

@@

## Cohort Curation and Sample Processing
We studied the differences in microbiome phage-bacteria interaction networks across healthy human bodies by leveraging previously published sequence sets containing purified virome samples paired with bacterial metagenomes from whole metagenomic shotgun sequences. Our study contained three datasets, including a study of the impact of diet on the healthy human gut virome [@Minot:2011ez], the impact of anatomical location on the healthy human skin virome [@Hannigan:2015fz], and the viromes of monozygotic twins and their mothers [@Reyes:2010cwa; @Turnbaugh:2009ei]. The viromes associated with these datasets were subjected to virus-like particle (VLP) purification to eliminate other organismal DNA including bacteria, fungi, and humans, and allowed us to assess the actively replicating virome.

The publications we utilized were conducted over the span of five years using different methods and technologies, and therefore yielded different sequence abundances **(Supplemental Figure \ref{SequenceStats} A-B)**. The bacterial and viral sequences from those publications were quality filtered and assembled into contigs. Because contig assembly most commonly returns genome fragments, we clustered related bacteria and phage contigs by k-mer frequency and co-abundance using the CONCOCT algorithm. These clusters represent operationally defined units of related bacteria and phage genomes that we defined as operational genomic units (OGUs). These OGUs are conceptually similar to the operational taxonomic unit (OTU) and operational protein family (OPF) definitions used for grouping highly similar 16S rRNA gene and open reading frame sequences, respectively [@Schloss:2008ht]. The resulting contigs and OGUs demonstrated high sequence coverage and length **(Supplemental Figure \ref{ContigStats} - \ref{ClusterStats})**.

## Establishing a Model of Phage-Bacteria Interactions
We predicted which phage OGUs infected which bacteria using a random forest model trained on experimentally validated infectious relationships from six previous publications [@Jensen:1998vh;@Malki:2015tm;@Schwarzer:2012ez;@Kim:2012dh;@Matsuzaki:1992gw;@Edwards:2015iz]. This training set contained diverse bacteria and phages, with both broad and specific infectious ranges **(Supplemental Figure \ref{ValidationOverview} A - B)**. Phages with linear and circular genomes, as well as ssDNA and dsDNA genomes, were included in the analysis. Because this was a DNA sequencing study, RNA phages were not considered in the analysis **(Supplemental Figure \ref{ValidationOverview} C-D)**. This training set included both positive relationships (a phage infects a bacterium) and negative relationships (a phage does not infect a bacterium). This built on previous work, which focused only on positive relationships, by allowing us to validate the false positive *and* false negative rates associated with our candidate models [@Edwards:2015iz].

Four phage and bacterial genomic markers were used to predict infectious relationships between bacteria and phages: 1) genome nucleotide similarities, 2) gene amino acid sequence similarities, 3) CRISPR targeting of phages by bacterial CRISPR spacer sequences, and 4) similarity of protein families known to be associated with known protein-protein interactions [@Orchard:2014hq]. The resulting random forest model performed well, with an AUC of `r signif(genmodelper$ROC, digits = sigfig)`, a sensitivity of `r signif(genmodelper$Sens, digits = sigfig)`, and a specificity of `r signif(genmodelper$Spec, digits = sigfig)` **(Figure \ref{RocCurve} A)**. The most important predictor in the model was nucleotide similarity between genes, followed by nucleotide similarity of whole genomes **(Figure \ref{RocCurve} B)**. Protein family interactions were moderately important to the model, while CRISPRs were minimally important, likely because they were redundant with the blast information. Approximately one third of the relationships yielded no score and were therefore unknown, unable to be assigned an interaction **(Figure \ref{RocCurve} C)**.

We used our random forest model to classify the relationships between bacteria and phages in the experimental datasets. The relationships within the three studies were used to construct one master network, containing the three study sub-networks, which themselves each contained sub-networks for each sample **(Figure \ref{RocCurve} D)**. Metadata including study, sample ID, disease, and abundance within the community (based on sequence count) were also stored in the multi-study master network to allow for effective parsing and downstream analysis **(Supplemental Figure \ref{NetworkDiagram})**. The resulting master network was highly connected and contained `r prettyNum(genfigurestats[c(genfigurestats$class %in% "Total"),"Edges"], big.mark = ",")` infectious relationships among `r prettyNum(genfigurestats[c(genfigurestats$class %in% "Total"),"Vertices"], big.mark = ",")` nodes, `r prettyNum(nestats[c(nestats$cats %in% "PhageNodes"),"values"], big.mark = ",")` of which represented phages and `r prettyNum(nestats[c(nestats$cats %in% "BacteriaNodes"),"values"], big.mark = ",")` that represented bacteria. Although the network was highly connected, not all relationships were present in all samples. Furthermore, relationships were weighted by relative abundance of their associated bacteria and phages, meaning lowly abundant relationships could be present but insignificant compared to those that were highly abundant. Like the master network, the skin network exhibited a diameter of `r prettyNum(genfigurestats[c(genfigurestats$class %in% "SkinStudy"),"Diameter"], big.mark = ",")` (measure of graph size; the greatest number of traversed vertices required between two vertices) and included almost all of the master network nodes and edges **(Figure \ref{RocCurve} E - F)**. The gut diet and twin sample sets each contained fewer than 150 vertices, fewer than 20,000 relationships, and diameters of `r prettyNum(genfigurestats[c(genfigurestats$class %in% "DietStudy"),"Diameter"], big.mark = ",")`, suggesting more sparsely related phages and bacteria **(Figure \ref{RocCurve} E - F)**.

## Individuality of Microbial Networks
Skin and gut community membership and diversity are highly personal, with people remaining more similar to themselves over time compared to other people [@Grice:2009ee; @Hannigan:2015fz; @Minot:2013ih]. We determined whether this personal conservation extended to network structure using our microbiome network model. We calculated the degree of dissimilarity between each subject's network, based on phage and bacteria abundance, as well as centrality. Centrality was evaluated by first calculating the weighted eigenvector centrality of all bacteria and phages within each sample graph. Conceptually, this metric defines central phages as those that are highly abundant and infect many bacteria which themselves are abundant and infected by many other phages. Bacterial centrality was defined using the same metric. We then calculated the similarity of community networks using the weighted eigenvector centrality of all nodes between all samples. Biologically, samples with similar network structures are interpreted as having similar capacities to influence other microbes within the community, transmit genetic material (e.g. horizontal gene transfer), and maintain stability.

We used this network dissimilarity metric to test whether microbiome network structures were more similar within people over time compared to other people. We found that gut microbiome network structure clustered by person (ANOSIM p-value = `r signif(interstats[c(interstats$site %in% "DietAnosim"), "prob"], digits = sigfig)`, R = `r signif(interstats[c(interstats$site %in% "AS"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} A**). We found evidence that network dissimilarity within people over the 8-10 day sampling period was less than the average dissimilarity between that person and others, although this was not statistically significant (p-value = `r signif(interstats[c(interstats$site %in% "Diet"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} B**). The lack of statistical confidence was likely due to the small sample size of this dataset. Although there was evidence for gut network conservation among individuals, we found no evidence for conservation of gut network structures within families. The gut network structures were not more similar between twins and their mothers (intrafamily) compared to other families of twins and mothers (inter-family) (p-value = `r signif(interstats[c(interstats$site %in% "Skin"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} C**).

Skin microbiome network structure was strongly conserved within individuals, and more adequately confirmed by our larger skin dataset (p-value = `r signif(interstats[c(interstats$site %in% "Twins"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} D**). This distribution was similar when separated by anatomical sites, with most sites also being significantly more conserved within individuals **(Supplemental Figure \ref{allskin})**.

## Role of Diet & Obesity in Gut Microbiome Conectivity
Diet is a major environmental factor that influences resource availability and gut microbiome composition and diversity, including bacteria and phages [@Minot:2011ez; @Turnbaugh:2009hf; @David:2014cl]. Previous work in isolated culture-based systems has suggested that changes in nutrient availability are associated with altered phage - bacteria network structures, although this has yet to be tested in humans [@Poisot:2011jc]. We hypothesized that a change in diet would be associated with a change in network structure.

We evaluated the diet-associated differences in gut network structure by quantifying how central each sample's network was on average across all microbes. We accomplished this by utilizing two common centrality metrics: degree centrality and closeness centrality. **Degree centrality**, the simplest centrality metric, was defined as the number of connections each phage made to bacteria, or each bacterium made to phages. Because this metric alone offers only minimal insight, we supplemented it with measurements of closeness centrality. **Closeness centrality** is a metric of how close each phage or bacterium is to all of the other phages and bacteria in the network. A higher closeness centrality suggests that genetic information or the effects of altered abundance would be more impactful to all other microbes in the system. A networks with higher closeness centrality also indicates an overall greater degree of connections, which suggests a greater resilience to instability. Because these values are assigned to each phage and bacterium within each network, we calculated the average connectedness and corrected for the maximum potential degree of connectedness to obtain a single value for the connectedness of each graph.

We found that gut microbiome network structures associated with high fat diets were less connected than those of low fat diets **(Figure \ref{dietnetworks} A-B)**. Tests for statistical differences were not performed due to the small sample size. High fat diets exhibited less degree centrality **(Figure \ref{dietnetworks} A)**. This suggests that bacteria were targeted by less phages and that phage tropism was more specific. High fat diets also exhibited decreased closeness centrality **(Figure \ref{dietnetworks} B)**, indicating that microbes were more distant from other microbes in the community. This would make genetic transfer and altered abundance less capable of impacting other bacteria and phages within the network.

In addition to diet, there was also an association between obesity and network structure **(Figure \ref{dietnetworks} C-D)**. The obesity-associated network demonstrated a higher degree centrality **(Figure \ref{dietnetworks} C)** but a less closeness centrality, compared to the healthy controls **(Figure \ref{dietnetworks} D)**. This suggested that the obesity network was overall less connected, having microbes further from all other microbes within the community.

## Variation of Network Structure Across the Human Skin Landscape
Extensive work has illustrated differences in healthy human skin microbiome between anatomical sites, including bacteria, viruses, and fungi [@Grice:2009ee; @Findley:2013jf; @Hannigan:2015fz]. These communities vary by degree of skin moisture, oil, and environmental exposure. We hypothesized that like microbial composition and diversity, microbial network structure differed between anatomical sites. We tested this hypothesis by evaluating the changes in network structure between anatomical sites within our skin dataset.

We quantified the average centrality of each sample again using the weighted eigenvector centrality metric. We found that intermittently moist skin sites (dynamic sites that fluctuate between being moist and dry) were significantly less connected than the more stable moist and sebaceous environments (p-value < 0.001, **Figure \ref{skinnetwork} A)**. We also found that skin sites that were protected from the environment (occluded) were much more highly connected than those that were constantly exposed to the environment or only intermittently occluded (p-value < 0.001, **Figure \ref{skinnetwork} B)**.

@@

# Discussion
We developed and implemented a network-based analytical approach that allowed us to begin evaluating the basic properties of the human microbiome through bacteria and phage *relationships*, instead of membership or diversity. The goal of this study was to provide an initial understanding of how phage-bacteria networks differ throughout the human body, so as to provide a baseline for future studies of how microbiome networks differ in disease states. This goal was similar to the initial studies of bacteria and viral diversity across the human body, as well as other environments [@Grice:2009eea; @Findley:2013jf; @Hannigan:2015fz; @Costello:2009im, @Consortium:2012iz; @Schloss:2005hz; @Minot:2011ez]. The use of network theory offered extensive analytical opportunities which could be applied to understand complex ecological communities. We focused on utilizing metrics of connectivity to understand the extent to which communities of bacteria and phages interact (e.g. horizontal gene transfer, modulated bacterial gene expression, alterations in abundance). By accomplishing this goal, we aimed to provide an initial understanding of human microbiome networks that will power fututure studies to use similar approaches in other setting such as disease states.
